Evolution of CSMA Protocols for the IEEE 802.11 Standard

### 2013-01-24 paper.tex
authors

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Michael Shell}
\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
Georgia Institute of Technology\\
Atlanta, Georgia 30332--0250\\
Email: http://www.michaelshell.org/contact.html}
\and
\IEEEauthorblockN{Homer Simpson}
\IEEEauthorblockA{Twentieth Century Fox\\
Springfield, USA\\
Email: homer@thesimpsons.com}
\and
\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
\IEEEauthorblockA{Starfleet Academy\\
San Francisco, California 96678-2391\\
Telephone: (800) 555--1212\\
Fax: (888) 555--1212}}

@@

A candidate to replace CSMA/CA should 
\begin{itemize}
  \item Provide performance advantages, either in the form of throughput or short term fairness.
  \item Be backward compatible with current implementation.
  \item Be simple a simple evolution implementation to ease the transition and reduce time to market (Optional but desirable).
\end{itemize}
\section{Related Work}
Since the popularization of IEEE 802.11, several papers have proposed modifications to the contention protocol that is used for sharing the medium.
They can be categorized in three groups regarding the approach they use.
The first one prevents that the contention window is reset to its minimum value after successful transmissions.
Examples of this first group include \cite{bharghavan1994map,wang2004ncr}.
This solution improves throughput in saturation conditions at the price of lowering short term fairness.

@@

Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
When a station joins the contention it initilalizes the retry attempt counter $r$ and the backoff stage $s$ to zero.
The backoff counter $b$ is initialized using a uniform random distribution and a contention window.
After each collision, the retry attempt counter and the backoff stage counter are incremented.
As a consecuence of the incremented backoff stage, a larger contention window is used.
Note that there is maximum backoff stage $S$ and a maximum retry limit $R$.
When the maximum retry  limit is reached the packet is discarded, $r$ and $s$ are reset anb a new value for $b$ is computed.
Similarly, if the packet is successfully transmitted, the $s$ and $r$ counters are reset and a new random backoff counter is computed.

@@

Hysteresis is included in \ref{alg:CSMA_ECA_hist}.
Adding hysteresis is as simple as removing lines 13 and 23 from Algoritm \ref{alg:CSMA_ECA}.

Finally, fair-share is implemented in \ref{alg:CSMA_ECA_hist_fair}.
Line 10 of \ref{alg:CSMA_ECA_hist} is modified to increas the number of transmitted packets from one to $2^s$.

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Luis Sanabria-Russo, Azadeh Faridi, Boris Bellalta, Jaume Barcelo, Miquel Oliver}
\IEEEauthorblockA{
Universitat Pompeu Fabra\\
Roc Boronat 183, 08018 Barcelona\\
Email: \{name.surname@upf.edu\}}
}
%\and
%\IEEEauthorblockN{Homer Simpson}
%\IEEEauthorblockA{Twentieth Century Fox\\
%Springfield, USA\\
%Email: homer@thesimpsons.com}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}}

@@

A candidate to replace CSMA/CA should 
\begin{itemize}
  \item Provide performance advantages, either in the form of throughput or short term fairness.
  \item Be backward compatible with current implementation.
  \item Support a large number of simultaneous contenders.
  \item Be simple a simple evolution implementation to ease the transition and reduce time to market (Optional but desirable).
\end{itemize}
\section{Related Work}
Since the popularization of IEEE 802.11, several papers have proposed modifications to the contention protocol that is used for sharing the medium.
They can be categorized in three groups regarding the approach they use.
The first one prevents that the contention window is reset to its minimum value after successful transmissions.
Examples of this first group include \cite{bharghavan1994map,wang2004ncr}.
This solution improves throughput in saturation conditions at the price of lowering short term fairness.

@@

In this section we describe the CSMA/CA protocol as it is currently implemented in the IEEE 802.11 and an evolution of the standard that satisfies the four requirements specified in the introduction.
Therefore, the presented protocol is a valid candidate to replace CSMA/CA in the upcoming revisions of the standard.

There are three changes in the new protocol compared to the legacy one.
Firstly, a deterministic backoff is used after successful transmissions.
Secondly, the backoff stage is not reset after a packet is serviced.
The backoff stage is reset only when the station leaves the contention because it has no packet to be transmitted.
And thirdly, the number of packets transmitted in every transmission attempt is a function of the backoff stage.
Note that current standards already support the transmission of multiple packets in a single slot.

In the following we describe the original protocol and then we describe each of the modifications that we propose, one by one.

Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
When a station joins the contention it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero.
The backoff counter $b$ is initialized using a uniform random distribution and a contention window.
After each collision, the retry attempt counter and the backoff stage counter are incremented.
As a consequence of the incremented backoff stage, a larger contention window is used.
Note that there is maximum backoff stage $S$ and a maximum retry limit $R$.
When the maximum retry  limit is reached the packet is discarded, $r$ and $s$ are reset anb a new value for $b$ is computed.
Similarly, if the packet is successfully transmitted, the $s$ and $r$ counters are reset and a new random backoff counter is computed.

@@

Hysteresis is included in \ref{alg:CSMA_ECA_hist}.
Adding hysteresis is as simple as removing lines 13 and 23 from Algorithm \ref{alg:CSMA_ECA}.
Note that when hysteresis is used, the backoff stage is reset only when the node has no packet to serve.

Finally, fair-share is implemented in \ref{alg:CSMA_ECA_hist_fair}.
Line 10 of \ref{alg:CSMA_ECA_hist} is modified to increase the number of transmitted packets from one to $2^s$.
